---
title: "Statistics_Genomics_coursera_JHU_week2"
author: "Asude Berber"
date: "2024-07-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
source("http://bioconductor.org/biocLite.R")
BiocManager::install("biocLite")
BiocManager::install("sva")
library(limma)
biocLite("devtools")  # only if devtools not yet installed
library(devtools)
library(Biobase)
```
##Dimension reduction 
#Can we find patterns in matrices of data?

```{r}
con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/montpick_eset.RData")
load(file=con)
close(con)
mp = montpick.eset
pdata=pData(mp)
edata=as.data.frame(exprs(mp))
fdata = fData(mp)
ls()
```

#calculate the singular vectors 
```{r}
edata = edata[rowMeans(edata) > 100, ]
edata = log2(edata + 1)
edata_centered = edata - rowMeans(edata)
svd1 = svd(edata_centered)
names(svd1)
```

```{r}
plot(svd1$d,ylab="Singular value",col=2)

# visualize the distribution of singular values obtained from the SVD. 
#If the singular values decrease rapidly and then level off, it indicates that the first few singular values are much larger than the rest, which can be important for tasks like dimensionality reduction.
#svd1: This is a variable that likely contains the result of a Singular Value Decomposition (SVD) operation. In R, svd1 would typically be a list containing at least three components: u, d, and v. Here, d represents the singular values from the decomposition.
#col=2 is for red

plot(svd1$d^2/sum(svd1$d^2),ylab="Percent Variance Explained",col=2)
```

```{r}
par(mfrow=c(1,2)) #this function is for displaying multiple plots in same area, comparing two plots 
plot(svd1$v[,1],col=2,ylab="1st PC") #variance
plot(svd1$v[,2],col=2,ylab="2nd PC")

```

#A very common plot is to plot PC1 versus PC2 to see if you can see any “clusters” or “groups”.
```{r}
plot(svd1$v[,1],svd1$v[,2],col=2,ylab="2nd PC",xlab="1st PC")

plot(svd1$v[,1],svd1$v[,2],ylab="2nd PC",
     xlab="1st PC",col=as.numeric(pdata$study))
```

```{r}
edata_outlier = edata_centered
edata_outlier[1,] = edata_centered[1,] * 10000
svd3 = svd(edata_outlier)
par(mfrow=c(1,2))
plot(svd1$v[,1],col=1,main="Without outlier")
plot(svd3$v[,1],col=2,main="With outlier")
```

#It turns out the new top singular vector is perfectly correlated with the outlying gene
```{r}
plot(svd3$v[,1],edata_outlier[1,],col=4)
```

#Reference: https://jtleek.com/genstats/inst/doc/02_03_dimension-reduction.html 


#Quiz questions
```{r}
#load the data 
con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/montpick_eset.RData")
load(file=con)
close(con)
mp = montpick.eset
pdata=pData(mp)
edata=as.data.frame(exprs(mp))
fdata = fData(mp)

#What percentage of variation is explained by the 1st principal component in the data set if you:

#Do no transformations?
pca_result = prcomp(edata, center = TRUE, scale. = TRUE)  # Standardize the data before PCA
# Variance explained by each principal component
pca_variance = pca_result$sdev^2

# Total variance
total_variance = sum(pca_variance)

# Percentage of variance explained by the first principal component
perc_variation_first_pc = (pca_variance[1] / total_variance) * 100

# Print the result
perc_variation_first_pc

```

```{r}
#with log2 transformation 
log_edata = log2(edata + 1)  # Apply log2 transformation
pca_result = prcomp(log_edata, center = TRUE, scale. = TRUE)  # Standardize the data before PCA
# Variance explained by each principal component
pca_variance = pca_result$sdev^2

# Total variance
total_variance = sum(pca_variance)

# Percentage of variance explained by the first principal component
perc_variation_first_pc = (pca_variance[1] / total_variance) * 100

# Print the result
perc_variation_first_pc

```

```{r}
#log2(data + 1) transform and subtract row means?
log_edata = log2(edata + 1)  # Apply log2 transformation
# Subtract the row means
log_edata_centered = t(apply(log_edata, 1, function(x) x - mean(x)))
pca_result = prcomp(log_edata_centered, center = FALSE, scale. = TRUE)  # Data is already centered, so set center = FALSE
# Variance explained by each principal component
pca_variance = pca_result$sdev^2

# Total variance
total_variance = sum(pca_variance)

# Percentage of variance explained by the first principal component
perc_variation_first_pc = (pca_variance[1] / total_variance) * 100

# Print the result
perc_variation_first_pc

```

## Question 2

#Load the Montgomery and Pickrell eSet:
#Perform the log2(data + 1) transform and subtract row means from the samples. Set the seed to \verb|333|333 and use k-means to cluster the samples into two clusters. Use \verb|svd|svd to calculate the singular vectors. What is the correlation between the first singular vector and the sample clustering indicator?

```{r}
con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/montpick_eset.RData")
load(file=con)
close(con)
mp = montpick.eset
pdata=pData(mp)
edata=as.data.frame(exprs(mp))
fdata = fData(mp)

set.seed(1235)

log_edata = log2(edata + 1)
log_edata_row_centered = log_edata - rowMeans(log_edata)

# Calculate SVD
e_svd = svd(log_edata_row_centered)

# Calculate kmeans
e_kmeans = kmeans(t(log_edata_row_centered), centers=2)

# Get correlation between 1st singular vector and cluster label
cor.test(e_svd$v[, 1], e_kmeans$cluster)

```

#Question3
```{r}
con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/bodymap_eset.RData")
load(file=con)
close(con)
bm = bodymap.eset
edata = exprs(bm)
pdata_bm=pData(bm)

#Fit a linear model relating the first gene’s counts to the number of technical replicates, treating the number of replicates as a factor. Plot the data for this gene versus the covariate.

# Extract the first gene's counts and number of technical replicates
first_gene_counts <- edata[1,]
num_replicates <- as.factor(pdata_bm$num.tech.reps)

lm_fit <- lm(first_gene_counts ~ num_replicates)
summary(lm_fit)

library(ggplot2)
plot(num_replicates, first_gene_counts,
     xlab = "Number of technical replicates",
     ylab = "First gene counts",
     main = "First gene counts vs number of technical replicates")
points(num_replicates, fitted(lm_fit), col = "red", pch = 19)

ggplot(lm_fit, aes(x = num_replicates, y = first_gene_counts)) +
  geom_point() +
  geom_point(aes(y = fitted(lm_fit)), color = "red") +
  labs(title = "First Gene Counts vs Number of Technical Replicates",
       x = "Number of Technical Replicates",
       y = "First Gene Counts")
```


#Question4
```{r}
con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/bodymap_eset.RData")
load(file=con)
close(con)
bm = bodymap.eset
edata = exprs(bm)
pdata_bm=pData(bm)

#Fit a linear model relating he first gene’s counts to the age of the person and the sex of the samples. What is the value and interpretation of the coefficient for age?

lm_fit2 <- lm(edata[1, ] ~ pdata_bm$age + pdata_bm$gender)
summary(lm_fit2)

#This coefficient means that for each additional year of age, the count goes down by an average of 23.91 for a fixed sex.
```

#Question5
```{r}
con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/montpick_eset.RData")
load(file=con)
close(con)
mp = montpick.eset
pdata=pData(mp)
edata=as.data.frame(exprs(mp))
fdata = fData(mp)

#Perform the log2(data + 1) transform. Then fit a regression model to each sample using population as the outcome. Do this using the 
#lm.fit
#lm.fit function (hint: don't forget the intercept). What is the dimension of the residual matrix, the effects matrix and the coefficients matrix?

edata_log2 <- log2(edata + 1)
population <- as.factor(pdata$population)

mod = model.matrix(~ pdata$population)
fit = lm.fit(mod, t(edata))

dim(fit$residuals)
dim(fit$effects)
dim(fit$coefficients)

?lm.fit
```


#Question7
```{r}
con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/bodymap_eset.RData")
load(file=con)
close(con)
bm = bodymap.eset
edata = exprs(bm)
pdata_bm=pData(bm)


#Fit many regression models to the expression data whereage age is the outcome variable using the  lmFit lmFit function from the  limma
#limma package (hint: you may have to subset the expression data to the samples without missing values of age to get the model to fit). What is the coefficient for age for the 1,000th gene? Make a plot of the data and fitted values for this gene. Does the model fit well?

library(limma)

# subset the expression data to the samples without mimssing values of age
pdata_bm = na.omit(pdata_bm)
edata = edata[,rownames(pdata_bm), drop=FALSE]

# fit many regression models to the expression data where age is the outcome
mod_adj = model.matrix(~ pdata_bm$age)
fit_limma = lmFit(edata,mod_adj)

fit_limma$coefficients[1000,]
```
```{r}
# make a plot of the 1,000th gene and fitted values
intercept = fit_limma$coefficients[1000,][1]
slope = fit_limma$coefficients[1000,][2]
x = edata[1000,]*slope+intercept

plot(x,pdata_bm$age)
```

#Question8 
```{r}
con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/bodymap_eset.RData")
load(file=con)
close(con)
bm = bodymap.eset
edata = exprs(bm)
pdata_bm=pData(bm)

#Fit many regression models to the expression data where age age is the outcome variable and tissue.type tissue.type is an adjustment variable using the lmFit lmFit function from the limma limma package (hint: you may have to subset the expression data to the samples without missing values of age to get the model to fit). What is wrong with this model?
pdata_bm$tissue.type

mod_adj = model.matrix(~ pdata_bm$tissue.type)
fit_limma = lmFit(edata,mod_adj)
fit_limma$coefficients[1000,]

head(pdata)
```
#The effects are difficult to distinguish because each study only measured one population. 

```{r}

con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/bodymap_eset.RData")
load(file=con)
close(con)
bm = bodymap.eset
edata = exprs(bm)
pdata_bm=pData(bm)
#Set the seed using the command 
set.seed(33353)
#then estimate a single surrogate variable using the 
#sva function 
#after log2(data + 1) transforming the expression data, removing rows with rowMeans less than 1, and treating age as the outcome (hint: you may have to subset the expression data to the samples without missing values of age to get the model to fit). What is the correlation between the estimated surrogate for batch and age? Is the surrogate more highly correlated with 
#race or gender?

library(sva)
pheno = na.omit(pdata_bm)
edata = edata[,rownames(pheno), drop=FALSE]
edata = log2(edata + 1)
edata = edata[rowMeans(edata) > 1,]




# fit a sva model
mod = model.matrix(~age, data=pheno)
mod0 = model.matrix(~1, data=pheno)
sva1 = sva(edata, mod,mod0, n.sv=2)


# correlation between surrogate for batch and age
cor(sva1$sv, pheno$age)
```
```{r}
# correlation between surrogate for batch and race
cor(sva1$sv, as.numeric(pheno$race))
```
```{r}
# correlation between surrogate for batch and gender
cor(sva1$sv, as.numeric(pheno$gender))
```
```


