---
title: "Statistics_Genomic_JHU_Coursera_Asude"
author: "Asude Berber"
date: "2024-07-03"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#R markdown lecture
#devtools dependencies
```{r}
??devtools
library(devtools)
par(pch=19)
?par #parameters

library(knitr)
knit_hooks$set(setPch = function(before, options, envir) {
  if(before) par(pch = 19)
})
opts_chunk$set(setPch = TRUE)

knitr::opts_chunk$set(fig.width=5, fig.height=5, size="footnotesize",
                      warning=FALSE, message=FALSE)
knitr::knit_hooks$set(small.mar = function(before, options, envir) {
  if (before) graphics::par(mar = c(5,5,1.5,1))
})

x = rnorm(100)
plot(x,col=3)

x = rnorm(100)
plot(x,col=3,pch=19)

?pch #plotting ‘character’
`````

#pch = 19: solid circle,

#pch = 20: bullet (smaller solid circle, 2/3 the size of 19),

#pch = 21: filled circle,

#pch = 22: filled square,

#pch = 23: filled diamond,

#pch = 24: filled triangle point-up,

#pch = 25: filled triangle point down.


##The three data tables in genomics
```{r}
library(devtools)
library(Biobase)

con=url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/bodymap_eset.RData")
load(file=con)
close(con)
bm = bodymap.eset
bm

expression_data = exprs(bm)
dim(expression_data)
head(expression_data)

pheno_data = pData(bm)
head(pheno_data)

feature_data = fData(bm)
dim(fData(bodymap.eset))

fData(bodymap.eset)[1:10, 1]
```
#Week1 Quiz Q&A
#for question2
```{r setup, eval=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

```{r }
x = rnorm(10)
plot(x,pch=19,col="dodgerblue")
```

```{r }
y = rbinom(20,size=1,prob=0.5)
table(y)
```

#for question3
```{r}
library(Biobase)
library(GenomicRanges)
data(sample.ExpressionSet, package = "Biobase")
se = makeSummarizedExperimentFromExpressionSet(sample.ExpressionSet)
library(SummarizedExperiment)

assay(se)
?SummarizedExperiment
colData(se)
rowRanges(se)
rowData(se)
pData(se)
```

#Just considering the phenotype data what are some reasons that the Bottomly data set is likely a better experimental design than the Bodymap data? Imagine the question of interest in the Bottomly data is to compare strains and in the Bodymap data it is to compare tissues.

```{r}
con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/bottomly_eset.RData")
load(file=con)
close(con)
bot = bottomly.eset
pdata_bot=pData(bot)

con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/bodymap_eset.RData")
load(file=con)
close(con)
bm = bodymap.eset
pdata_bm=pData(bm)
```

#What are some reasons why this plot is not useful for comparing the number of technical replicates by tissue (you may need to install the plotrix package).
```{r}
con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/bodymap_eset.RData")
load(file=con)
close(con)
bm = bodymap.eset
pdata_bm=pData(bm)

library(plotrix)
pie3D(pdata_bm$num.tech.reps,labels=pdata_bm$tissue.type)
```

#Question7 
#Which of the following code chunks will make a heatmap of the 500 most highly expressed genes (as defined by total count), without re-ordering due to clustering? Are the highly expressed samples next to each other in sample order?
```{r}
con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/bodymap_eset.RData")
load(file=con)
close(con)
bm = bodymap.eset
edata = exprs(bm)

row_sums = rowSums(edata)
edata = edata[order(-row_sums),]
index = 1:500
heatmap(edata[index,],Rowv=NA,Colv=NA)

#without re-ordering?
row_sums = rowSums(edata)
#index = which(rank(row_sums) < 500)
#heatmap(edata[index,],Colv=NA)

index = which(rank(-row_sums) < 500 )
heatmap(edata[index,],Rowv=NA,Colv=NA)
```

#Question8
#Make an MA-plot of the first sample versus the second sample using the log2 transform (hint: you may have to add 1 first) and the rlog rlog transform from the DESeq2 package. How are the two MA-plots different? Which kind of genes appear most different in each plot?
```{r}
con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/bodymap_eset.RData")
load(file=con)
close(con)
bm = bodymap.eset
pdata = pData(bm)
edata = exprs(bm)

#A very widely used plot is what is known as a M-A plot, sometimes called a Bland Altman plot. The basic idea is to plot the sum of the two values on the x-axis and the difference on the y-axis. This can be used to see any difference between the (samples, averages, etc.) and to see if there is any intensity-specific biases.

mm = log2(edata[,1]+1) - log2(edata[,2]+1)
aa = log2(edata[,1]+1) + log2(edata[,2]+1)
plot(aa,mm,col=2)
```
```{r}
library(DESeq2)
rld <- rlog(exprs(bm))

y_rld = rld[,1] - rld[,2]
x_rld = rld[,1] - rld[,2]
plot(x_rld, y_rld, col = "blue", type = "p")
```


##Question9 
```{r}
con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/montpick_eset.RData")
load(file=con)
close(con)
mp = montpick.eset
pdata=pData(mp)
edata=as.data.frame(exprs(mp))
fdata = fData(mp)
```

Cluster the data in three ways:

1. With no changes to the data
2. After filtering all genes with rowMeans less than 100
3. After taking the log2 transform of the data without filtering
Color the samples by which study they came from (Hint: consider using the function myplclust.R in the package rafalib available from CRAN and looking at the argument lab.col.)

How do the methods compare in terms of how well they cluster the data by study? Why do you think that is?

```{r}
#with no changes
# By default calculates the distance between rows
dist1 = dist(t(edata))

## Look at distance matrix
colramp = colorRampPalette(c(3,"white",2))(9)
heatmap(as.matrix(dist1),col=colramp,Colv=NA,Rowv=NA)

hclust1 = hclust(dist1)
plot(hclust1)
plot(hclust1,hang=-1)

#after filtering 
low_genes = rowMeans(edata) < 100
filter_edata = filter(edata, !low_genes)
f_dist1 = dist(t(filter_edata))
f_hclust1 = hclust(f_dist1)

#par(mar=c(0, 4, 4, 2))
plot(f_hclust1, hang = -1, main="remove low expression", labels=FALSE)

#After taking the log2 transform of the data without filtering
log_edata = log2(edata + 1)
l_dist1 = dist(t(log_edata))
l_hclust1 = hclust(l_dist1)

par(mar=c(0, 4, 4, 2))
plot(l_hclust1, hang=-1, main="perform log2 transform", labels=FALSE)
```


#Question10
```{r}
con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/montpick_eset.RData")
load(file=con)
close(con)
mp = montpick.eset
pdata=pData(mp)
edata=as.data.frame(exprs(mp))
fdata = fData(mp)
```
Cluster the samples using k-means clustering after applying the 
log2
log2 transform (be sure to add 1). Set a seed for reproducible results (use 
set.seed(1235)
set.seed(1235)). If you choose two clusters, do you get the same two clusters as you get if you use the 
cutree
cutree function to cluster the samples into two groups? Which cluster matches most closely to the study labels?

```{r}
edata = log2(edata + 1)
#kmeans
set.seed(1235)
kmeans1 = kmeans(edata,centers=2)
names(kmeans1)
matplot(t(kmeans1$centers),col=1:3,type="l",lwd=3)
```

```{r}
dist1 = dist(t(edata))
hclust1 = hclust(dist1)
tree = cutree(hclust1, 2)

par(mar=c(0, 4, 4, 2))
plot(hclust1, tree, main="cutree")
```

##They produce different answers, with hierarchical clustering giving a much more unbalanced clustering. The k-means clustering matches study better. 
