---
title: "Statistics_Week3_Genomic_JHU_Coursera"
author: "Asude Berber"
date: "2024-07-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
install.packages(c("devtools","broom","MASS"))
BiocManager::install(c("snpStats"))
library(devtools)
library(Biobase)
library(snpStats)
library(broom)
library(MASS)
library(DESeq2)
```


#echo=FALSE

##Generalized linear models in R
```{r}
#Data from SNP , from a case-control genome wide association study. 
data(for.exercise)
use <- seq(1, ncol(snps.10), 10)
sub.10 <- snps.10[,use]

#calculate PCs 
xxmat <- xxt(sub.10, correct.for.missing = FALSE)
evv <- eigen(xxmat, symmetric=TRUE)
pcs <- evv$vectors[,1:5]

#First we do an unadjusted logistic regression assuming an additive model.The coefficient is the change in log-odds for a one unit decrease (because homozygous major allele is coded 1) in the number of copies of the minor allele.
snpdata = sub.10@.Data
status = subject.support$cc
snp1 = as.numeric(snpdata[,1])
snp1[snp1==0] = NA
glm1 = glm(status ~ snp1,family="binomial")
tidy(glm1)

#For example suppose we want to code a dominant model (so only an association of risk with the two copies of the common allele, now the coefficient on snp1_dom is the increase in log odds associated with two copies of the major allele).

snp1_dom = (snp1 == 1)
glm1_dom = glm(status ~ snp1_dom,family="binomial")
tidy(glm1_dom)

tidy(glm1)

glm2 = glm(status ~ snp1 + pcs[,1:5],family="binomial")
tidy(glm2)
```

#for question 1 of the quiz
##Fit a linear model and a logistic regression model to the data for the 3rd SNP. What are the coefficients for the SNP variable? How are they interpreted? (Hint: Don't forget to recode the 0 values to NA for the SNP data)
```{r}
snp3 = as.numeric(snpdata[,3])
snp3[snp3==0] = NA
glm3 = glm(status ~ snp3,family="binomial")
tidy(glm3)
#logistic model = -0.16 

lm3 <- lm(status ~ snp3)
tidy(lm3)

#linear model = -0.04
```
##logistic regression is better choice than linear regression because 
#If you included more variables it would be possible to get negative estimates for the probability of being a case from the linear model, but this would be prevented with the logistic regression model. 



##Question3
##Fit a logistic regression model on a recessive (need 2 copies of minor allele to confer risk) and additive scale for the 10th SNP. Make a table of the fitted values versus the case/control status. Does one model fit better than the other?
```{r}
#with the same data set
snp10 = as.numeric(snpdata[,10])
snp10[snp10==0] = NA
glm10 = glm(status ~ snp10,family="binomial")
tidy(glm10)

snp10_dom = (snp10 == 2)
glm10_dom = glm(status ~ snp10_dom,family="binomial")
tidy(glm10_dom)
```

#Question4
##### Fit an additive logistic regression model to each SNP. What is the average effect size? What is the max? What is the minimum?
```{r}
# fit an additive logistic regression model to each SNP
results = rep(NA, dim(snpdata)[2])
for (i in 1:ncol(snpdata)){
  snpdata_i = as.numeric(snpdata[,i])
  snpdata_i[snpdata_i == 0] = NA
  glm_i = glm(status ~ snpdata_i, family = "binomial")
  results[i] = tidy(glm_i)$statistic[2]
}

# average effect size
mean(results)
```
```{r}
# minimum effect size
min(results)
```
```{r}
# maximum effect size
max(results)
```


#Question5
#Fit an additive logistic regression model to each SNP and square the coefficients. What is the correlation with the results from using snp.rhs.tests 
#and chi.squared? 
#Why does this make sense?
```{r}
#For logistic regression modeling of many SNPs at once we can use the snps.rhs.tests function which computes an asymptotic chi-squared statistic. This isn’t quite the same thing as the F-statistics we have been calculating but can be used in the same way for significance calculations.
glm_all = snp.rhs.tests(status ~ 1,snp.data=sub.10)
slotNames(glm_all)
qq.chisq(chi.squared(glm_all),df=1)

# square the coefficients
results_coeff_squre =  results^2

cor(results_coeff_squre, chi.squared(glm_all))
```

##Question6
```{r}
con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/montpick_eset.RData")
load(file=con)
close(con)
mp = montpick.eset
pdata=pData(mp)
edata=as.data.frame(exprs(mp))
fdata = fData(mp)

#Do the log2(data + 1) transform and fit calculate F-statistics for the difference between studies/populations using genefilter:rowFtests and using genefilter:rowttests. Do you get the same statistic? Do you get the same p-value?

edata_log2 <- log2(edata + 1)

if (!requireNamespace("genefilter", quietly = TRUE)) {
  install.packages("genefilter")
}
library(genefilter)

# Ensure that the grouping variable is a factor
group_factor = factor(pdata$study)  # Replace 'study' with the actual column name representing studies/populations

log_edata_matrix = as.matrix(edata_log2)

# Calculate F-statistics
ftest_results = rowFtests(log_edata_matrix, group_factor)

# Calculate T-statistics (assuming two groups for t-test)
# Ensure that the grouping variable has exactly two levels for t-test
if (length(levels(group_factor)) == 2) {
  ttest_results = rowttests(log_edata_matrix, group_factor)
} else {
  stop("rowttests is only applicable for two-group comparisons.")
}

# Extract F-statistics and p-values
f_stat = ftest_results$statistic
f_pvalue = ftest_results$p.value

# Extract T-statistics and p-values (for two groups)
t_stat = ttest_results$statistic
t_pvalue = ttest_results$p.value

# Compare
comparison = data.frame(F_Statistic = f_stat, F_P_Value = f_pvalue,
                        T_Statistic = t_stat, T_P_Value = t_pvalue)

print(head(comparison))


```
#You get the same p-value but different statistics. This is because the F-statistic and t-statistic test the same thing when doing a two group test and one is a transform of the other. 


#Question7
```{r}
con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/montpick_eset.RData")
load(file=con)
close(con)
mp = montpick.eset
pdata=pData(mp)
edata=as.data.frame(exprs(mp))
edata = edata[rowMeans(edata) > 100,]
fdata = fData(mp)

#First test for differences between the studies using the DESeq2 package using the DESeq function.

# using DESeq2 test the differences between the studies
de = DESeqDataSetFromMatrix(edata, pdata, ~study)
glm_de = DESeq(de)
result_de = results(glm_de)
result_de
```

```{r}
# Then do the log2(data + 1) transform
# Apply log2 transformation
log_edata <- log2(edata + 1)

# Create design matrix
design <- model.matrix(~ pdata$study)

#do the test for differences between studies using the 
#limma package and the lmFit, ebayesand topTable functions. 
#What is the correlation in the statistics between the two analyses? Are there more differences for the large statistics or the small statistics (hint: Make an MA-plot).

# Fit linear model
fit <- lmFit(log_edata, design)
fit <- eBayes(fit)
res_limma <- topTable(fit, coef = 2, number = Inf, sort.by = "none")

# Extract statistics
deseq2_stat <- result_de$stat
limma_stat <- res_limma$t

# Calculate correlation
correlation <- cor(deseq2_stat, limma_stat, use = "complete.obs")
print(correlation)

```
```{r}
# MA-plot for DESeq2
plotMA(result_de, main="DESeq2 MA-plot")

# MA-plot for limma
plot(rowMeans(log_edata), res_limma$logFC, main="limma MA-plot", xlab="Mean Expression(log2 scale)", ylab="log Fold Change")

```

#Question8: Apply the Benjamni-Hochberg correction to the P-values from the two previous analyses. How many results are statistically significant at an FDR of 0.05 in each analysis?
```{r}
fp_bh = p.adjust(result_de$pvalue ,method="BH")
sum(fp_bh < 0.05)

fp_bh = p.adjust(res_limma$P.Value ,method="BH")
sum(fp_bh < 0.05)
```


