---
title: "Bioconductor_Week3_Genomic_DataScience_JHU_Asude"
author: "Asude Berber"
date: "2024-06-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(Biobase)
library(ALL)
library(hgu95av2.db)

if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("minfiData")

#source("http://www.bioconductor.org/biocLite.R")
#biocLite(c("Biobasne", "ALL", "hgu95av2.db"))
```


#ALL package is experimental data package 
```{r}
data(ALL)
ALL
experimentData(ALL)

#the exprs matrix containing the 12625 gene expression measurements on the 128 samples (a 12625 by 128 numeric matrix).
#the pData data.frame containing phenotype data on the samples.

exprs(ALL)[1:4, 1:4]
head(sampleNames(ALL))
head(featureNames(ALL))
head(pData(ALL))
head(ALL$sex)

```

```{r}
#Question 1
#Question: What is the mean expression across all features for sample 5 in the ALL dataset (from the ALL package)?
mean(exprs(ALL)[,5])

```


```{r}
#the SummarizedExperiment class from GenomicRanges package; an extension of the ExpressionSet class to include GRanges.
#example dataset
library(airway)
#airway package downloaded
data(airway)
airway
```
```{r}
#Question6: 
#We are using the airway dataset from the airway package.
#Question: What is the average of the average length across the samples in the expriment?
length(rowRanges(airway))
library(GenomicRanges)
mean(airway$avgLength)


#Question7 We are using the airway dataset from the airway package. The features in this dataset are Ensembl genes.
#Question: What is the number of Ensembl genes which have a count of 1 read or more in sample SRR1039512?
sum(assay(airway)[,3]>=1)

#Question8
#Question: The airway dataset contains more than 64k features. How many of these features overlaps with transcripts on the autosomes (chromosomes 1-22) as represented by the TxDb.Hsapiens.UCSC.hg19.knownGene package?
#Clarification: A feature has to overlap the actual transcript, not the intron of a transcript. So you will need to make sure that the transcript representation does not contain introns.

library(GenomicFeatures)
library(TxDb.Hsapiens.UCSC.hg19.knownGene)

# exon data of txdb
txdb <- TxDb.Hsapiens.UCSC.hg19.knownGene
txdb_exons <- exons(TxDb.Hsapiens.UCSC.hg19.knownGene)

# transcripts on the autosome
autosome <- paste0("chr", c(1:22))
txdb_exons_autosome <- keepSeqlevels(txdb_exons, autosome, pruning.mode = "coarse")

# rename in NCBI format
txdb_ncbi <- mapSeqlevels(seqlevels(txdb_exons), "NCBI")
txdb_exons_ncbi <- renameSeqlevels(txdb_exons_autosome, txdb_ncbi)

dim(subsetByOverlaps(airway, txdb_exons_ncbi))[1]

#Question9
 #The expression measures of the airway dataset are the number of reads mapping to each feature. In the previous question we have established that many of these features do not overlap autosomal transcripts from the TxDb.Hsapiens.UCSC.hg19.knownGene. But how many reads map to features which overlaps these transcripts?  
sample_SRR1039508 <- airway[, 1]
sample_SRR1039508_autosome <- subsetByOverlaps(sample_SRR1039508, txdb_exons_ncbi)
sample_SRR1039508_autosome

# Question: For sample SRR1039508, how big a percentage (expressed as a number between 0 and 1) of the total reads in the airway dataset for that sample, are part of a feature which overlaps an autosomal TxDb.Hsapiens.UCSC.hg19.knownGene transcript?
autosome_reads <- sum(assay(sample_SRR1039508_autosome, "counts"))
total_reads <- sum(assay(sample_SRR1039508, "counts"))

# percentage of the total reads in airway dataset for SRR1039508 which overlaps autosome of txdb
autosome_reads/total_reads
```




```{r}

library(GEOquery)
library(biomaRt)


#Biomart” is a flexible interface to a biological database. The idea is that any kind of resource can setup a Biomart and then users can access the data using a single set of tools to access multiple databases.
```
#Question2
```{r}
#We will use the biomaRt package to annotate an Affymetrix microarray. We want our results in the hg19 build of the human genome and we therefore need to connect to Ensembl 75 which is the latest release on this genome version. How to connect to older versions of Ensembl is described in the biomaRt package vignette; it can be achived with the command

mart <- useMart(host='feb2014.archive.ensembl.org', biomart = "ENSEMBL_MART_ENSEMBL")

#Question: Using this version of Ensembl, annotate each feature of the ALL dataset with the Ensembl gene id. How many probesets (features) are annotated with more than one Ensembl gene id?
library("hgu95av2.db")
ensembl <- useDataset("hsapiens_gene_ensembl", mart)

# annotate each feature
feature_name <- featureNames(ALL)
annotation_ALL <- getBM(attributes=c("ensembl_gene_id","affy_hg_u95av2"), filters="affy_hg_u95av2", values=feature_name, mart=ensembl)

sum(table(annotation_ALL[,2])>1)

#3.Question 3
#Question: How many probesets (Affymetrix IDs) are annotated with one or more genes on the autosomes (chromosomes 1 to 22).
# list Attributes
attributes <- listAttributes(ensembl)
filters <- listFilters(ensembl)

# annotate autosomes
chrom <- c(1:22)
annotation_ALL_chr <- getBM(attributes=c("ensembl_gene_id", "affy_hg_u95av2", "chromosome_name"), filters=c("affy_hg_u95av2","chromosome_name"), values=list(feature_name, chrom), mart=ensembl)

sum(table(table(annotation_ALL_chr[,2])))
```
##Question4
#Use the MsetEx dataset from the minfiData package. Part of this question is to use the help system to figure out how to address the question.  
#Question: What is the mean value of the Methylation channel across the features for sample “5723646052_R04C01”?

```{r}
library(minfiData)
library(minfi)
View(MsetEx)
MsetEx #An example dataset for Illumina's Human Methylation 450k dataset, after preprocessing.
mean(getMeth(MsetEx)[,2]) #coldata name second sample
```

#Question 5
#Question: Access the processed data from NCBI GEO Accession number GSE788. What is the mean expression level of sample GSM9024?
```{r}
eList <- getGEO("GSE788")
class(eList)
eData <- eList[[1]]
eData
mean(exprs(eData)[,2])
```

#Question10
#Consider sample SRR1039508 and only consider features which overlap autosomal transcripts from TxDb.Hsapiens.UCSC.hg19.knownGene. We should be able to very roughly divide these transcripts into expressed and non expressed transcript. Expressed transcripts should be marked by H3K4me3 at their promoter. The airway dataset have assayed “airway smooth muscle cells”. In the Roadmap Epigenomics data set, the E096 is supposed to be “lung”. Obtain the H3K4me3 narrowPeaks from the E096 sample using the AnnotationHub package.  
```{r}
library(AnnotationHub)
ah <- AnnotationHub()
ah_E096 <- query(ah, c("E096", "H3K4me3", "narrowPeak"))
ah_record <- ah_E096[["AH30596"]]

#ah_record_autosome <- keepSeqlevels(ah_record, autosome, pruning.mode = "coarse")
#ah_record_ncbi <- renameSeqlevels(ah_record_autosome, txdb_ncbi)

#ncbi_group <- extractSeqlevelsByGroup(species = "Homo sapiens", style = "NCBI", group = "auto")
#sample_ncbi <- keepSeqlevels(range(rowRanges(sample_SRR1039508_autosome)), ncbi_group)

#ov <- subsetByOverlaps(promoters(sample_ncbi), ah_record_ncbi)
#ov <- subsetByOverlaps(sample_SRR1039508, ov)

#median(assay(ov, "counts"))

#reference: https://github.com/miotomato/Coursera_Genomic_Data_Science_Specialization/blob/master/Genomic_Bioconductor/Bioconductor_quiz3.Rmd 
```

